Legacy Data Confounds Modern Genomics Studies

### 2018-05-15 legacy.tex
implemented some of alex's comments

\newcommand{\sgcomment}[1]{\textcolor{blue}{SG: #1}}
\newcommand{\luke}[1]{\textcolor{blue}{Luke: #1}}
\newcommand{\todo}[1]{\textcolor{blue}{*#1*}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Legacy Data Confounds Modern Genomics Studies}

@@

Many of these large datasets have been assembled over many years. 
The first sequenced genomes in the 1000 Genomes project (1kGP), for example, were sequenced \todo{10} years ago, at a time when the accuracy of sequencing technologies was still being ascertained. %\sgcomment{blah blah blah}. 
Because of the extraordinary value of freely available data, this early data from the 1kGP is still widely used as a reference panel for imputation, allele frequency estimations and a wide range of applications.

@@

%A genome wide mutational signature can be measured by taking the sum of all the different types of mutations in an individual. 
%\sgcomment{This is a procedural definition. You need a conceptual definition! Something like:} 
Mutations can occur in the genome as a result of environmental inflictions as well as genetic factors like DNA replication and repair \cite{}.
Many mutagens leave signatures that are specific to a genomic context \cite{}.
This means that the neighbouring nucleotides of a mutation can be used to gain insight into the cause of a mutation \cite{}.
Thymine dimers caused by UV radiation were detected early on because they were highly specific and easy to study \cite.
However,  signatures caused by natural variations in the sequence of proteins involved in proofreading and repair are more difficult to study because they tend to occur less frequently \cite{}.
A large sample size can be used to resolve some of these fine scale mutational signatures \cite{}.
%Finer scale patterns have been reported in \todo{MORE HERE} \cite{}.

In 2015, Harris et al reported a \todo{15\%} overrepresentation of the TCC to TTC mutation in European populations.
Harris and Pritchard published a follow up study comparing how various ethnic groups differ from one another with respect to their mutational spectrum. 
In this paper, they proposed that not only is there a distinct mutational spectrum difference in Europeans, but that these changes are present in most human populations. 
One result was of particular interest the heterogeneously distributed signal found in a subset of Japanese individuals. 
A mutational signature that is present unevenly across a population is unexpected because it suggested strong population structure; unlikely for a dense population living on an archipelago.

@@

Datasets are often combined to increase power or diversity of a dataset.
However, combining data can be an issue especially when the quality of the data produced can vary between sequencing technologies and even sequencing centres.
The errors in one dataset don't disappear when they are combined with another higher quality one.
This is rapidly becoming a relevant issue as newer technologies have lower error rates, and quality controls become more stringent.
Many of the legacy datasets produced using dated sequencing technologies have been known to contain a higher rate of false positives than their more recent counterparts.

@@

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{Figure1.jpg}
\caption{
\textbf{A} 
Joint frequency spectrum plot of the Japanese from the 1000 Genomes Project and a more recent dataset.
Crosses ( + ) are variants that reached genome wide significance in a GWA of sequencing quality. 
The histogram on the left of the plot is the distribution of significant variants. 
The lowest frequency variant able to reach genome wide significance is 5\%.
\textbf{B} 
Mutation spectrum of the 1048 variants that reached a genome wide significance with a p value less than  in a in a GWA of sequencing quality. 
There is a \todo{significant} enrichment in *AC${\rightarrow}$*CC mutations. 
\textbf{C} 
Genome wide association of the average quality of mapped bases for the 104 Japanese individuals included in the 1000 Genomes Project. This GWA study identified $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual
The same analysis was performed independently for each of the populations in the 1000 Genomes Project. }
 \label{SFS}
\end{figure}

@@

Despite removing the variants identified as being low quality, the signal persists. 
When we remove the variants significantly associated to low quality, the enrichment in *AC${\rightarrow}$*CC mutations remains present, but \todo{less} significant. 
When we remove individuals with average quality of mapped SNPs below 30, the signal goes away. 
We suspect that this has to do with the lack of power we have for more rare alleles. 
The lowest frequency variants that are significantly associated to quality are at 5\%. 
When we remove all the variants above 1\% in 1kGP but missing from the NAG data, the signal disappears.
This suggests that while the GWA approach can identify some of the low hanging bad apples, there are likely more of these false positives nested inside these legacy cohorts.

@@

As more and more large scale genotyping efforts are being imputed on the same legacy datasets, we must scrutinize the quality of the reference databases to avoid the propagation of false positives. 
These results bring forth many questions regarding the reliability of legacy datasets. 
Moreover, since there are so many broad applications of imputation, it frames the question for reference data turnover.

>>>>

\newcommand{\sgcomment}[1]{\textcolor{blue}{SG: #1}}
\newcommand{\luke}[1]{\textcolor{blue}{Luke: #1}}
\newcommand{\todo}[1]{\textcolor{blue}{*#1*}}
\newcommand{\alex}[1]{\textcolor{red}{Alex: #1}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Legacy Data Confounds Modern Genomics Studies}

@@

Many of these large datasets have been assembled over many years. 
The first sequenced genomes in the 1000 Genomes project (1kGP), for example, were sequenced \todo{10} years ago, at a time when sequencing technologies were still being . %\sgcomment{blah blah blah}. 
Because of the extraordinary value of freely available data, this early data from the 1kGP is still widely used as a reference panel for imputation, allele frequency estimations and a wide range of applications.

@@

%A genome wide mutational signature can be measured by taking the sum of all the different types of mutations in an individual. 
%\sgcomment{This is a procedural definition. You need a conceptual definition! Something like:} 
Mutations can occur in the genome as a result of environmental inflictions as well as variation in genetic factors like DNA proofreading and repair \cite{}.
Many mutagens leave signatures that are specific to a genomic context \cite{}.
This means that the neighbouring nucleotides of a mutation can be used to gain insight into the cause of a mutation \cite{}.
Thymine dimers caused by UV radiation were detected early on because they were highly specific and easy to study \cite.
However,  signatures caused by natural variations in the sequence of proteins involved in proofreading and repair are more difficult to study because they tend to occur less frequently and are more subtle \cite{}.
A large sample size can be used to increase statistical power to resolve some of these fine scale mutational signatures \cite{}.
%Finer scale patterns have been reported in \todo{MORE HERE} \cite{}.

In 2015, Harris et al reported a \todo{15\%} overrepresentation of the TCC to TTC mutation in European populations.
Harris and Pritchard published a follow up study comparing how various ethnic groups differ from one another with respect to their mutational spectrum. 
In this paper, they proposed that not only is there a distinct mutational spectrum difference in Europeans, but that many populations have their own mutational signatures.
One result was of particular interest the heterogeneously distributed signal found in a subset of Japanese individuals. 
A mutational signature that is present unevenly across a population is unexpected because it suggested strong population structure; unlikely for a dense population living on an archipelago.

@@

Many types of DNA sequencing machines have been available on the market over the past decade.
Different types of sequencing technologies can have different technical limitations in quality control.
This makes combining data an issue especially when multiple sequencing technologies are involved in the data production.
Many of the legacy datasets produced using dated sequencing technologies have been known to contain a higher rate of false positives than their more recent counterparts.
The errors in one dataset do not disappear when they are combined with another higher quality one.
The technical biases caused by legacy sequencing technology is becoming increasingly relevant as newer technologies produce data with lower error rates, and more stringent quality controls.

@@

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{Figure1.jpg}
\caption{
\textbf{A} 
Joint frequency spectrum plot of the Japanese from the 1000 Genomes Project and a more recent dataset.
Crosses ( + ) are variants that reached genome wide significance in a GWA of sequencing quality. 
The histogram on the left of the plot is the distribution of significant variants. 
The lowest frequency variant able to reach genome wide significance is 5\%.
\textbf{B} 
Mutation spectrum of the 1048 variants that reached a genome wide significance with a p value less than in a GWA of sequencing quality. 
There is a \todo{significant} enrichment in *AC${\rightarrow}$*CC mutations. 
\textbf{C} 
Genome wide association of the average quality of mapped bases for the 104 Japanese individuals included in the 1000 Genomes Project. This GWA study identified $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual
The same analysis was performed independently for each of the populations in the 1000 Genomes Project. }
 \label{SFS}
\end{figure}

@@

\alex{Having trouble following the first three sentences}
Despite removing the variants identified as being low quality, the signal persists. 
When we remove the variants significantly associated to low quality, the enrichment in *AC${\rightarrow}$*CC mutations remains present, but \todo{less} significant. 
When we remove individuals with average quality of mapped SNPs below 30, the signal goes away. 
We suspect that this has to do with the lack of power we have for more rare alleles. 
The lowest frequency variants that are significantly associated to quality are at 5\%. 
When we remove all the variants above 1\% in 1kGP but missing from the NAG data, the signal disappears.
This suggests that while the GWA approach can identify some of the low hanging bad apples, there are likely more of these false positives nested inside these legacy cohorts.

@@

As more large scale genotyping efforts are being imputed on the same legacy datasets, we must scrutinize the quality of the reference databases to avoid the propagation of false positives. 
These results bring forth many questions regarding the reliability of legacy datasets. 
Moreover, since there are so many broad applications of imputation, it frames the question for reference data turnover.
